Let's talk about Thurstone & Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2024-11-01 index.qmd
section:introduction,issues,structural; process:write

<!-- 1. What does CJ do? -->
[In *comparative judgment* (CJ) studies, judges assess the presence of a trait or competence by conducting pairwise comparisons of stimuli [@Thurstone_1927; @Pollitt_2004; @Pollitt_2012a]. The comparison produces a dichotomous outcome, indicating which stimulus is perceived to possess a higher level of the trait. After conducting multiple rounds of pairwise comparisons, researchers use the Bradley-Terry-Luce (BTL) model [@Bradley_et_al_1952; @Luce_1959] to process the outcomes and estimate scores that reflect the underlying trait of interest. CJ has been successfully employed in assessing the quality of written texts [@Laming_2004; @Pollitt_2012b; @Whitehouse_2012; @vanDaal_et_al_2016; @Lesterhuis_2018; @Coertjens_et_al_2017; @Goossens_et_al_2018; @Bouwer_et_al_2023].]{style="color:red;"}

@@

<!-- 2. Where are we now with DCJ? -->
Numerous studies have documented the effectiveness of CJ in assessing various traits and competencies over the past decade. These studies have emphasized three aspects of the method's effectiveness: its reliability, validity, and practical applicability. Research on reliability indicates that CJ requires a relatively small number of pairwise comparisons [@Verhavert_et_al_2019; @Crompvoets_et_al_2022] to produce trait scores that are as precise and consistent as those generated by other assessment methods [@Coertjens_et_al_2017; @Goossens_et_al_2018; @Bouwer_et_al_2023]. Furthermore, evidence suggests that the reliability and time efficiency of CJ are comparable, if not superior, to those of other assessment methods when employing adaptive comparison algorithms [@Pollitt_2012b; @Verhavert_et_al_2022; @Mikhailiuk_et_al_2021]. On the other hand, research on validity suggests that scores generated by CJ can accurately represent the traits being measured [@Whitehouse_2012; @vanDaal_et_al_2016; @Lesterhuis_2018; @Bartholomew_et_al_2018; @Bouwer_et_al_2023]. Finally, research on practical applicability highlights the method's versatility across both educational and non-educational contexts [@Jones_2015; @Bartholomew_et_al_2018; @Jones_et_al_2019; @Marshall_et_al_2020; @Bartholomew_et_al_2020; @Boonen_et_al_2020].

@@

<!-- 3. But what are the problems of DCJ as a method? -->
Nevertheless, despite the growing number of studies on CJ, unsystematic and fragmented research approaches in the literature have left several critical issues unaddressed. This research primarily focuses on three: the apparent disconnect between CJ's measurement and structural model, the over-reliance on the assumptions of Thurstone's Case 5 [-@Thurstone_1927] in CJ's measurement model, and the unclear role of comparison algorithms on the method's reliability and validity. The following sections will discuss each of these issues in detail, followed by the introduction of a theoretical model and its statistical translation, which aim to address all three concerns simultaneously.

@@

<!-- 2. But there is no structural model, only post estimation analyses -->
Researchers then typically use the estimated BTL scores, or their transformations, to conduct additional analyses or hypothesis testing. The literature indicates that these scores have been used to identify 'misfit' judges and stimuli [@Pollitt_2012b; @vanDaal_et_al_2017; @Goossens_et_al_2018], detect biases in judges' ratings [@Pollitt_et_al_2003; @Pollitt_2012b], calculate correlations with other scoring methods [@Goossens_et_al_2018; @Bouwer_et_al_2023], or test hypotheses related to the latent trait of interest [@Bramley_et_al_2019; @Boonen_et_al_2020; @Bouwer_et_al_2023; @vanDaal_et_al_2017; @Jones_et_al_2019; @Gijsen_et_al_2021].

<!-- 3. But what is the problem with this? -->
However, the statistical literature cautions against using estimated scores to conduct additional analyses or tests. A key consideration is that BTL scores are parameter estimates that inherently carry uncertainty. Ignoring this uncertainty when conducting additional analyses and tests can inflate their precision and statistical power, increasing the risk of committing a type I error [@McElreath_2020], which is when a null hypothesis is incorrectly rejected [@Everitt_et_al_2010].

<!-- 4. So, what is the solution?  -->
To mitigate these risks, principles from Structural Equation Modeling (SEM) [@Hoyle_et_al_2023; @Kline_et_al_2023] and Item Response Theory (IRT) [@deAyala_2009; @Fox_2010; @vanderLinden_et_al_2017] recommend conducting these analyses and tests within a structural model that accounts for both the scores and their uncertainties, rather than treating them separately. Thus, an integrated approach combining CJ's structural and measurement models can offer significant advantages.

>>>>

<!-- 1. What does CJ do? -->
[In *comparative judgment* (CJ) studies, judges assess the presence of a trait or competence by conducting pairwise comparisons of stimuli [@Thurstone_1927; @Pollitt_2004; @Pollitt_2012a]. The comparison produces a dichotomous outcome, indicating which stimulus is perceived to possess a higher trait level. After conducting multiple rounds of pairwise comparisons, researchers use the Bradley-Terry-Luce (BTL) model [@Bradley_et_al_1952; @Luce_1959] to process the outcomes and estimate scores that reflect the underlying trait of interest. This method has been successfully employed in assessing the quality of written texts [@Laming_2004; @Pollitt_2012b; @Whitehouse_2012; @vanDaal_et_al_2016; @Lesterhuis_2018; @Coertjens_et_al_2017; @Goossens_et_al_2018; @Bouwer_et_al_2023].]{style="color:red;"}

@@

<!-- 2. Where are we now with DCJ? -->
Numerous studies have documented the effectiveness of CJ in assessing various traits and competencies over the past decade. These studies have emphasized three aspects of the method's effectiveness: its reliability, validity, and practical applicability. Research on reliability indicates that CJ requires a relatively small number of pairwise comparisons [@Verhavert_et_al_2019; @Crompvoets_et_al_2022] to produce trait scores that are as precise and consistent as those generated by other assessment methods [@Coertjens_et_al_2017; @Goossens_et_al_2018; @Bouwer_et_al_2023]. Furthermore, evidence suggests that the reliability and time efficiency of CJ are comparable, if not superior, to those of other assessment methods when employing adaptive comparison algorithms [@Pollitt_2012b; @Verhavert_et_al_2022; @Mikhailiuk_et_al_2021]. Additionally, research on validity suggests that scores generated by CJ can accurately represent the traits under measurement [@Whitehouse_2012; @vanDaal_et_al_2016; @Lesterhuis_2018; @Bartholomew_et_al_2018; @Bouwer_et_al_2023], while research on practical applicability highlights the method's versatility across both educational and non-educational contexts [@Jones_2015; @Bartholomew_et_al_2018; @Jones_et_al_2019; @Marshall_et_al_2020; @Bartholomew_et_al_2020; @Boonen_et_al_2020].

@@

<!-- 3. But what are the problems of DCJ as a method? -->
Nevertheless, despite the growing number of CJ studies, unsystematic and fragmented research approaches in the literature have left several critical issues unaddressed. This research primarily focuses on three: the apparent disconnect between CJ's measurement and structural model, the over-reliance on the assumptions of Thurstone's Case 5 [-@Thurstone_1927] in CJ's measurement model, and the unclear role of comparison algorithms on the method's reliability and validity. The following sections will discuss each of these issues in detail, followed by the introduction of a theoretical model and its statistical translation, which aim to address all three concerns simultaneously.

@@

<!-- 2. But there is no structural model, only post estimation analyses -->
Researchers then typically use the estimated BTL scores, or their transformations, to conduct additional analyses or hypothesis testing. The literature indicates that the scores have been used to identify 'misfit' judges and stimuli [@Pollitt_2012b; @vanDaal_et_al_2017; @Goossens_et_al_2018], detect biases in judges' ratings [@Pollitt_et_al_2003; @Pollitt_2012b], calculate correlations with other assessment methods [@Goossens_et_al_2018; @Bouwer_et_al_2023], or test hypotheses related to the underlying trait of interest [@Bramley_et_al_2019; @Boonen_et_al_2020; @Bouwer_et_al_2023; @vanDaal_et_al_2017; @Jones_et_al_2019; @Gijsen_et_al_2021].

<!-- 3. But what is the problem with this? -->
However, the statistical literature cautions against using estimated scores to conduct additional analyses or tests. A key consideration is that BTL scores are parameter estimates that inherently carry uncertainty. Ignoring this uncertainty when conducting separate analyses and tests can inflate their precision and statistical power, increasing the risk of committing a type I error [@McElreath_2020], which is when a null hypothesis is incorrectly rejected [@Everitt_et_al_2010].

<!-- 4. So, what is the solution?  -->
To mitigate these risks, principles from Structural Equation Modeling (SEM) [@Hoyle_et_al_2023; @Kline_et_al_2023] and Item Response Theory (IRT) [@deAyala_2009; @Fox_2010; @vanderLinden_et_al_2017] recommend conducting these analyses and tests within a structural model.  A structural model specify how different manifest or latent variables influence the latent variable of interest [@Everitt_et_al_2010]. This approach allows analyses that can account for both the scores and their uncertainties simultaneously, rather than treating them as separate elements. Therefore, an integrated approach that combines CJ’s structural and measurement models can offer significant advantages.
